Topic 6.2: Open Data

### 2020-04-15 draft_chapter.Rmd
added some literature

---
title: "Topic 6.2: Open Data"
author: "Samuel Langton (MMU) & Reka Solymosi (University of Manchester)"
date: "June 2020"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

@@

## Introduction

- Topic definition.
- Advantages and exciting prospects in open data.
- Chapter structure outline.
- Prerequisites (reference _Geocomputation with R_).

## Background

- What is open data?
- What types of organisations release open data?
  - Public sector
  - Private sector
- Why is it needed?
- What are the strengths?
- What are the limitations?
- Substantive examples in criminology (focus on crime and place research).
- Conclude with: a key one is Open Street Map.

@@

To demonstrate the data in its entirety, we can plot the Greater London boundaries, overlayed with the bicycle parking space locations, and open crime data about bicycle thefts. It is worth clarifying that open police recorded crime data in England and Wales is spatially anonymised by a process of snapping points to a pre-defined grid (see Tompson et al., 2015). For that reason, many of these points overlap, and thus a degree of transparency is used for the points.

@@

@@

## References

Tompson, L., Johnson, S., Ashby, M., Perkins, C., & Edwards, P. (2015). UK open source crime data: accuracy and possibilities for research. _Cartography and geographic information science_, 42(2), 97-111.

>>>>

---
title: "Topic 6.2: Open Data"
author: "Samuel Langton (MMU) & Reka Solymosi (University of Manchester)"
date: "June 2020"
output: pdf_document
editor_options: 
  chunk_output_type: console
bibliography: refs.bib
---

@@

# Introduction

Access to most data sets have traditionally been restricted to those who have the means to collect these themselves or to pay for, or otherwise gain access to them. Even when access is granted, this is often with conditions that circumscribe how the data can be used through licensing or policy [@kitchin2014data]. This lead to unequal access to data and the subsequent insight that can be attained from them.

Open data are a response to these restrictions, seeking to open up data without restrictions, to broaden access and participation in research deriving insight from all sorts of data sets, removing the need to ask permission or negotiate access [@manovich2011trending]. Not only that, but using open data can lead to all sorts of novel insight within the domain of criminology and crime analysis specifically by tapping into constructs and processes difficult to capture through surveys, interviews, and other traditional measures [@solymosi2018role]. Therefore, it is important that social scientists, researchers, crime analysts, and others interested in making sense of the social world around them have the skills and know-how to access and accurately interpret, assess and analyse such data sets.

This chapter aims to lay the groundwork for developing such skills, providing both a framework to approach and meainingfully interpret open data regarding understanding its origin and any limitations associated with that, and a practical hands-on guide to accessing, wrangling, and analysting open data to draw conclusions about crime and place. We achieve this by first giving a background on open data, discussing the key types of open data out there, critically assessing the strengths and limiations of open data, and building a how-to guide checklist for researchers to follow when approaching an open data source. Finally, we work through an example that shows how to access, wrangle, link, and interpret open data from various sources, providing a template to follow.

- Prerequisites (reference _Geocomputation with R_).

# Background

## What is open data?

What do we mean when we refer to **open data**? According to @dietrich2009open in the Open Data Handbook, open data is "data that can be freely used, re-used and redistributed by anyone - subject only, at most, to the requirement to attribute and sharealike."

They summarise the content of the Open Definition document published by the @knowledge2016open into three key tenants of open data:

- **Availability and Access:** the data must be available in the publica domain, accessible at no more than a reasonable reproduction cost. It should be in a machine readable form that is easily accessed and modified. For example, it should be downloadable as a spreadsheet, or json file, rather than presented as a summary table in a PDF document. 
- **Re-use and Redistribution:** the data must be provided under terms that allow free use, including modification, separation, or compilation of the data. permit re-use and redistribution including the intermixing with other datasets.
- **Universal Participation:** everyone must be able to use, re-use and redistribute the data and its derivatives without restriction (eg: 'non-commercial' restrictions or use in education purposes only)

---
In short, open data should be data that is available in a useful format for analysis that can be used by anyone for any purpose. 
---

Open data practices vary within and between countires. For example, the United States of America have a history of making public sector data sets available, while the United Kingdom is much more strict in releasing their data under a licence or for a fee [@kitchin2014data]. The Global Open Data Index [@globopendata] is one tool for tracking the annual global benchmark for publication of open government data by countries, providing a rank which shows how well governments across the world follow open data practices.

## What are types of open data?

Here we formulate a typology of open data sets based on their origin.

### Public sector (including publicly funded research)

Much of the open data movement has focused on data generated by state agencies (often termed public sector information) or publicly funded research, given that these have been funded by the public purse for the public's benefit  [@kitchin2014data]. For example, the Global Open Data Index introduced earlier focuses specifically on governmental bodies' data publishers, identifying data gaps in these organisations, and promoting them to think about how to make their data more useable and eventually more impactful [@globopendata]. This is important as open public sector data can be ....

<!-- MORE HERE basically the paragraph above should really say what is public sector data, why its important (ie what can we do with it as researhcers) -->

Regarding opening up research data, the argument focuses not only on the need to make open data where its collection was funded by public funds, but also on the need for transparency and reproducibility. Replication is a hallmark of such open science practice, and opening up data sets used in publications in criminology and wider social sciences contributes to foster a culture of replication [@pridemore2018replication]. Open data and replications fit into the wider discourse in social sciences around open data, open review and open dialogue, which may even become (with time) policy requirements regardless of the research funding source [@vuong2017open]. There are repositories where open research data is available (osf.io, uk data service, others??).

<!-- Somehow I want to also say "keep in mind open research practices when collecting your own data, although we dont cover how to actually release and make open your own data in this chapter, read more about this somewhere... idk? -->

### Private sector

A more limited focus of open data activists lays on  opening up data generated by private industry, which might have more proprietary value to its creators [@kitchin2014data]. While datasets are released by private sector companies, they often are only a subset selected by the organisation. However they might still provide valuable insight. Indeed many interesting papers have been written using the data released by large private sector organisations such as Google [@stephens2017everybody], Instagram [@reece2017instagram], Uber [@davidson2017interactive] and many more. In another chapter in this book you will learn to use the free Twitter API to do just the same (see Nick and Dan's chapter link to that).

### Open crowdsourced data

Finally, there is a specific group of open data sources that collate data collectively generated by large groups of individuals who do not specifically belong to any organisation, but instead work together on a collaborative project. These are crowdsourced data sets. Examples include Wikipedia (www.wikipedia.org), an online encyclopedia where anyone can contribute, and Flickr (www.flickr.com) an online photo gallery where anyone can upload and tag their photos. Often the organisations who maintain and monitor these data collection activities are charities or non govermental organisations that operate not for profit, but instead to provide some sort of social good. An example is the online problem reportin platform fixmystreet, where people report instances of graffiti or other environmental complaints, which can offer insight into people's experiences with incivilities [@solymosi2018crowdsourcing]. Another chapter in this book (REF CHAPTER WITH DAVID) discusses the merits and pitfalls of crowdsourced data, and what to watch out for when analysing open data of this type.

## What are strengths and limitations

Each one of the above categories might have its own strengths and limitations, and it is very important to consdier the source of the data you are working with, and what might mean for what and who is represented or excluded from these data. Here we will provide a general overview, and in the next section some guidance on how to make sure you take all necessary precautions when analysing open data of any kind.

Open data is hailed for its ease of use afforded by its availability and access. This strength is emphasised throughout the chapter. However another advantage and motivation for using open data is its potential to address many limitations associated with traditional surveying methods, such as social desirability bias, or issues associated with memory and recall [@mayer2013big].

However this also creates a limitation for such data, in that because it is already available, the researcher has no say into what gets collected, and instead must make do with what is available. There are limits as to what questions researchers can ask of a data set, and what interpretations are appropriate [@boyd2012critical].

Despite these limitation, the social sciences must embrace these new forms of data that, although messy, biased and noisy, have the potential to describe social phenomena better than well-organized small surveys or even national censuses [@savage2007coming].

## What do researchers need to know to make good use of open data

in the case of open data, there is no need to ask for permissions or negotiate data access in order to use it for research [@manovich2011trending]. In this sense, the great advantage of this data is that it is already out there. To collect it, all that is needed is a way to be able to interpret what this data means (for example by applying a framework for its analysis), and some skill in data scraping, wrangling and cleaning, in order to be able to transform it into a usable format for research [@boyd2012critical].

necessary to ask critical questions about what all this data means, who gets access to what data, how data analysis is deployed, and to what ends [@boyd2012critical].

be careful in interpreting people's communication online as authentic [@manovich2011trending].

The key concerns about these dara relate to issues about sampling, and ultimately the generalisability of any findings that can be drawn from them. In the case of crowdsourced data, where the sample of the population who contributes to such data is self-selected, giving way for people more motivated to speak about the issue, which means the data are likely to contain inherent bias [@longley2012geodemographics]. Specifically it tends to be men, between ages of 20-50, with a college or university degree who are most likely contributors [@budhathoki2010participants; @haklay2010good].

Looking into what contextual factors influence participation in Open Street Map, @mashhadi2013putting find that, socio-economic factors such as population density, dynamic population, distance from the centre and poverty all play an important role. These are important to keep in mind when reporting findings based on analysis of such data.

even with this limitation, the inclusion of crowdsourced data which are both up-to-date and specific to the problem at hand can still provide new insight in addition to the knowledge from established sources of data collected in traditional methods [@birkin2011calibration]. All of this helps us to understand the relations between the events and the occurrences that come together in unique places and provides a framework for understanding how places change over different time periods [@lonlongley2012geodemographics].

@@

To demonstrate the data in its entirety, we can plot the Greater London boundaries, overlayed with the bicycle parking space locations, and open crime data about bicycle thefts. It is worth clarifying that open police recorded crime data in England and Wales is spatially anonymised by a process of snapping points to a pre-defined grid (see @tompson2015uk). For that reason, many of these points overlap, and thus a degree of transparency is used for the points.

@@

### Bus stop example

```{r}
bus_sf <- opq(bbox = bb_sf) %>%                                 # select bounding box
  add_osm_feature(key = 'highway', value = 'bus_stop') %>%      # select features
  osmdata_sf() %>%                                              # specify class
  trim_osmdata(bb_poly = bb_sf)  
```

@@

# References
